---
title: "RESPOND v1 vs v2 End-to-End Benchmark"
format:
  html:
    toc: true
execute:
  echo: false
---

```{r}
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)
```

## Overview

This report measures the **total wall-clock runtime** of the full RESPOND
simulation pipeline for v1 and v2, covering three phases:

| Phase | v1 | v2 (respondpy) |
|---|---|---|
| **Load** | Read CSV input files → R matrices | Read SQLite DB → `Simulation` object |
| **Run** | Rcpp `sim()` | `sim.run()` |
| **Write** | `write.table()` to CSV | Polars → CSV files |

### Scaling methodology

Each "sample" represents one independent pipeline execution:

- **v2**: one cohort ID in the SQLite database. `N` samples means `N` cohort
  rows (all backed by the same parameter tables); `build_simulation()` loads
  and runs all `N` cohorts in a single call.
- **v1**: one full read → sim → write cycle. `N` samples means the entire
  pipeline is repeated `N` times with fixed-size inputs (13 intervention
  blocks, 1 age group, 1 sex group, 4 OUD states, 52 steps).

Because the two models scale differently by design—v2 horizontally (more
cohort instances) and v1 vertically via repeated pipeline calls—direct
per-sample comparisons reflect architectural throughput rather than identical
workloads.

---

## Inputs

```{r}
combined_path <- "data/combined_e2e_runtime.csv"
if (!file.exists(combined_path)) {
    stop(
        "Missing data/combined_e2e_runtime.csv.",
        " Run scripts/run_v1_e2e_benchmark.R and scripts/run_v2_e2e_benchmark.py,",
        " then scripts/prepare_e2e_benchmark_data.R."
    )
}

runtime <- read_csv(combined_path, show_col_types = FALSE) |>
    mutate(
        model = factor(model, levels = c("v1", "v2")),
        scope = factor(scope, levels = c("cold_start", "steady_state")),
        phase = factor(phase, levels = c("combined", "load", "run", "write"))
    )

required_models <- c("v1", "v2")
present_models <- unique(as.character(runtime$model))
if (!all(required_models %in% present_models)) {
    stop("Combined e2e runtime file must include both models: v1 and v2.")
}

required_phases <- c("combined", "load", "run", "write")
present_phases <- unique(as.character(runtime$phase))
if (!all(required_phases %in% present_phases)) {
    stop(
        sprintf(
            "Missing phases: %s",
            paste(setdiff(required_phases, present_phases), collapse = ", ")
        )
    )
}

required_scopes <- c("cold_start", "steady_state")
present_scopes <- unique(as.character(runtime$scope))
if (!all(required_scopes %in% present_scopes)) {
    stop(
        sprintf(
            "Missing scopes: %s",
            paste(setdiff(required_scopes, present_scopes), collapse = ", ")
        )
    )
}
```

---

## Combined Pipeline Summary

```{r}
combined_runtime <- runtime |>
    filter(scope == "steady_state", phase == "combined")

summary_tbl <- combined_runtime |>
    group_by(model) |>
    summarise(
        mean_of_means_job_ms = mean(mean_job_ms, na.rm = TRUE),
        median_of_means_job_ms = median(mean_job_ms, na.rm = TRUE),
        mean_p95_job_ms = mean(p95_job_ms, na.rm = TRUE),
        mean_std_job_ms = mean(std_job_ms, na.rm = TRUE),
        mean_of_means_sample_ms = mean(mean_sample_ms, na.rm = TRUE),
        n_sample_sizes = n_distinct(sample_size),
        .groups = "drop"
    ) |>
    arrange(model)

kable(summary_tbl,
    digits = 3,
    caption = "Combined pipeline (load + run + write) summary across all sample sizes"
)
```

```{r}
compare_tbl <- combined_runtime |>
    select(model, sample_size, mean_job_ms, p95_job_ms, std_job_ms, mean_sample_ms) |>
    pivot_wider(
        names_from  = model,
        values_from = c(mean_job_ms, p95_job_ms, std_job_ms, mean_sample_ms),
        names_glue  = "{.value}_{model}"
    ) |>
    mutate(
        mean_job_ratio_v1_over_v2 = mean_job_ms_v1 / mean_job_ms_v2,
        p95_job_ratio_v1_over_v2 = p95_job_ms_v1 / p95_job_ms_v2,
        mean_sample_ratio_v1_over_v2 = mean_sample_ms_v1 / mean_sample_ms_v2
    ) |>
    arrange(sample_size) |>
    select(-c(p95_job_ms_v1, p95_job_ms_v2, std_job_ms_v1, std_job_ms_v2, mean_sample_ms_v1, mean_sample_ms_v2, mean_job_ratio_v1_over_v2, p95_job_ratio_v1_over_v2, mean_sample_ratio_v1_over_v2))

kable(compare_tbl,
    digits = 3,
    caption = "Steady-state combined runtime: v1 vs v2 comparison by sample size"
)
```

---

## Combined Pipeline Runtime vs Sample Size

```{r fig.width=8, fig.height=4.5}
ggplot(combined_runtime, aes(x = sample_size, y = mean_job_ms, color = model)) +
    geom_line(linewidth = 0.9) +
    geom_point(size = 2.5) +
    geom_errorbar(
        aes(ymin = mean_job_ms - std_job_ms, ymax = mean_job_ms + std_job_ms),
        width = 0.5, alpha = 0.5
    ) +
    scale_x_continuous(breaks = sort(unique(combined_runtime$sample_size))) +
    labs(
        title  = "Steady-state end-to-end runtime: combined pipeline",
        x      = "Sample size (N pipeline iterations / cohorts)",
        y      = "Mean job runtime (ms)",
        color  = "Model"
    ) +
    theme_bw()
```

---

## User Throughput Comparison

Throughput is expressed as **samples processed per second**, derived from the
steady-state per-sample cost (`1000 / mean_sample_ms`). This normalizes v1's
repeated-pipeline runs and v2's cohort-batch runs to a common "samples/sec"
unit for direct comparison.

```{r}
throughput_tbl <- combined_runtime |>
    mutate(throughput_samples_per_sec = 1000 / mean_sample_ms) |>
    select(model, sample_size, mean_sample_ms, throughput_samples_per_sec)

speedup_tbl <- throughput_tbl |>
    pivot_wider(
        id_cols = sample_size,
        names_from = model,
        values_from = throughput_samples_per_sec,
        names_glue = "throughput_{model}"
    ) |>
    mutate(speedup_v2_over_v1 = throughput_v2 / throughput_v1) |>
    arrange(sample_size)

kable(speedup_tbl,
    digits = 2,
    caption = "v2 throughput speedup over v1 by sample size"
)
```

```{r fig.width=8, fig.height=4.5}
ggplot(throughput_tbl, aes(x = sample_size, y = throughput_samples_per_sec, color = model)) +
    geom_line(linewidth = 0.9) +
    geom_point(size = 2.5) +
    scale_x_continuous(breaks = sort(unique(throughput_tbl$sample_size))) +
    scale_y_log10() +
    scale_color_manual(values = c(v1 = "#f0325f", v2 = "#3d9be9")) +
    labs(
        title = "Log-Scaled Throughput Comparison",
        x     = "Sample Size (N iterations / cohorts)",
        y     = "Log-Scaled Throughput (samples/sec)",
        color = "Model"
    ) +
    theme_bw()
```

```{r fig.width=8, fig.height=4.5}
ggplot(speedup_tbl, aes(x = factor(sample_size), y = speedup_v2_over_v1)) +
    geom_col(fill = "#3d9be9") +
    geom_hline(yintercept = 1, linetype = "dashed", color = "#f0325f") +
    geom_text(aes(label = sprintf("%.1fx", speedup_v2_over_v1)), vjust = -0.4, color = "#003771") +
    labs(
        title = "v2 throughput speedup over v1",
        x     = "Sample size",
        y     = "Speedup (v2 throughput / v1 throughput)"
    ) +
    theme_bw()
```

```{r fig.width=8, fig.height=4.5}
ggplot(combined_runtime, aes(x = sample_size, y = mean_sample_ms, color = model)) +
    geom_line(linewidth = 0.9) +
    geom_point(size = 2.5) +
    scale_x_log10(breaks = sort(unique(combined_runtime$sample_size))) +
    scale_y_log10() +
    labs(
        title = "Per-sample cost vs. sample size (log-log)",
        x     = "Sample size (log scale)",
        y     = "Mean time per sample (ms, log scale)",
        color = "Model"
    ) +
    theme_bw()
```

---

## Phase Breakdown

```{r}
phase_tbl <- runtime |>
    filter(scope == "steady_state") |>
    group_by(model, phase) |>
    summarise(
        mean_of_means_job_ms = mean(mean_job_ms, na.rm = TRUE),
        mean_p95_job_ms = mean(p95_job_ms, na.rm = TRUE),
        .groups = "drop"
    ) |>
    arrange(phase, model)

kable(phase_tbl,
    digits = 3,
    caption = "Mean runtime (ms) per phase, averaged across all sample sizes"
)
```

```{r fig.width=9, fig.height=5}
ggplot(runtime |> filter(scope == "steady_state"), aes(x = sample_size, y = mean_job_ms, color = model)) +
    geom_line(linewidth = 0.8) +
    geom_point(size = 2) +
    facet_wrap(~phase, scales = "free_y", nrow = 2) +
    scale_x_continuous(breaks = sort(unique(runtime$sample_size))) +
    labs(
        title  = "Steady-state job runtime per phase vs sample size",
        x      = "Sample size",
        y      = "Mean job runtime (ms)",
        color  = "Model"
    ) +
    theme_bw()
```

---

## Phase Composition (v2 only)

For v2 the three phases are measured separately, showing how load, run, and
write each contribute to the total pipeline cost.

```{r fig.width=7, fig.height=4.5}
v2_phases <- runtime |>
    filter(scope == "steady_state", model == "v2", phase != "combined") |>
    group_by(phase, sample_size) |>
    summarise(mean_job_ms = mean(mean_job_ms, na.rm = TRUE), .groups = "drop")

ggplot(v2_phases, aes(x = sample_size, y = mean_job_ms, fill = phase)) +
    geom_col(position = "stack") +
    scale_x_continuous(breaks = sort(unique(v2_phases$sample_size))) +
    labs(
        title = "v2 pipeline phase composition by sample size",
        x     = "Sample size (cohort count)",
        y     = "Stacked mean job runtime (ms)",
        fill  = "Phase"
    ) +
    theme_bw()
```

---

## Notes

- **Warmup**: Each benchmark script discards an initial set of warmup
  iterations before timing begins (default: 3).
- **Repetitions**: Timing statistics are computed across multiple repetitions
  (default: 3), with all per-sample times pooled.
- **Sample sizes**: `1 5 10 25 50` (smaller grid than the simulation-only
  benchmark to keep total run time reasonable given the repeated I/O cost).
- **v1 cost analysis**: disabled (sentinel `-1` values) so the simulation
  code path matches the parity configuration used for the simulation-only
  benchmark.
